Inference for multivariate peaks over threshold models

### 2022-01-05 methodology_projgamma.tex
update

A key feature of the the projected Gamma distribution is its computational properties. We augment  $\mathcal{PG}(\bm{y}_i\mid\bm{\alpha}_i,\bm{\beta}_i) $ by introducing a latent radial component $r_i$, for each observation. Using Equation \eqref{joint} we observe that the 
full conditional of $r_i$ is easy to sample from, as it is given as
\begin{equation}
    r_i\mid\bm{\alpha}_i,\bm{\beta}_i \sim \mathcal{G}\left(\sum_{\ell = 1}^d\alpha_{i\ell}, \sum_{\ell = 1}^d\beta_{\ell} y_{i\ell}\right).
\end{equation}
Moreover,  the full conditional for $\bm{\alpha}_j,\bm{\beta}_j$ is then proportional to
\begin{equation}
    f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\bm{\alpha}_j,\bm{\beta}_j).
\end{equation}
We first consider a centering distribution given by a product of independent Gammas:
\begin{equation*}
    G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}).
\end{equation*}
This model is completed with independent Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.  We also assume a Gamma prior on $\eta$, that is updated via the procedure outlined in \cite{escobar1995}.  We refer to this model as the \emph{projected Gamma--Gamma} (PG--G) model.  An advantage of the PG--G model is that, thanks to conjugacy, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.  Then, the full conditional for $\alpha_{j\ell}$ takes the form
\begin{equation*}
    \pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}\times\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\Gamma^{n_j}(\alpha_{j\ell})\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
\end{equation*}
for $\ell = 2,\ldots,d$.  For $\ell = 1$, as $\beta_{1} := 1$, the full conditional takes the form
\begin{equation}
    \label{eqn:alpha1update}
    \pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\zeta_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
\end{equation}
Samples of $\alpha_{j\ell}$ can thus be obtained using a Metropolis step. In our analysis, we first transform $\alpha_{j\ell}$ to the log scale, and use a normal proposal density.
The full conditional for $\beta$ is 
\begin{equation}
    \beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \sim \mathcal{G}\left(\beta_{j\ell}\mid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
\end{equation}
for $\ell = 2,\ldots, d$.  Updating $\beta_{j\ell}$ is done via a Gibbs step.  The hyper-parameters $\xi_{\ell},\tau_{\ell},\zeta_{\ell},\sigma_{\ell}$ follow similar Gamma-Gamma update relationships.  We also explore a restricted form of this model, where $\beta_{\ell} = 1$ for all $\ell$.  Under this model, we use the full conditional in Equation~\ref{eqn:alpha1update} for all $\ell$, and omit inference on $\bm{\zeta},\bm{\sigma}$.  We refer to this model as the \emph{projected restricted Gamma--Gamma} (PRG--G) model.

The second form of centering distribution we explore is a multivariate log-normal distribution on the shape parameters $\bm{\alpha}_j$, with independent Gamma $\beta_{j\ell}$ rate parameters.  
\begin{equation}
    G_0\left(\bm{\alpha}_j,\bm{\beta}_j\mid\bm{\mu},\bm{\Sigma},\zeta,\sigma\right) = \mathcal{LN}\left(\bm{\alpha}_j\mid\bm{\mu},\bm{\Sigma}\right)\times\prod_{\ell = 2}^d\mathcal{G}\left(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}\right).
\end{equation}
This model is completed with a normal prior on $\bm{\mu}$, an inverse Wishart prior on $\bm{\Sigma}$, and Gamma priors on $\zeta_{\ell}$ and $\sigma_{\ell}$, and $\eta$.  This model is denoted as the \emph{projected Gamma--log-normal} (PG--LN) model.  We also explore a restricted Gamma form of this model as above, where $\beta_{\ell} = 1$ for all $\ell$.  This is denoted as the \emph{projected restricted Gamma--log-normal} (PRG--LN) model.  In the interest of brevity, we omit the inferential forms for this model, sufficing to say that the Gamma prior for $\beta_{j\ell}$ means it can again be integrated out, improving mixing on sampling $\bm{\alpha}_j$ with a joint Metropolis step; the normal prior for $\mu$ is conjugate for the log-normal $\bm{\alpha}_j$, and sampled via a Gibbs step; and an inverse Wishart prior for $\Sigma$ is again conjugate to the log-normal $\bm{\alpha}_j$, implying that it can also be sampled via a Gibbs step.

>>>>

A key feature of the the projected Gamma distribution is its computational properties. We augment  $\mathcal{PG}(\bm{y}_i\mid\bm{\alpha}_i,\bm{\beta}_i) $ by introducing a latent radial component $r_i$, for each observation. Using Equation \eqref{joint} we observe that the 
full conditional of $r_i$ is easy to sample from, as it is given as
\begin{equation}
    r_i\mid\bm{\alpha}_i,\bm{\beta}_i \sim \mathcal{G}\left(\sum_{\ell = 1}^d\alpha_{i\ell}, \sum_{\ell = 1}^d\beta_{\ell} y_{i\ell}\right).
\end{equation}
Moreover,  the full conditional for $\bm{\alpha}_j,\bm{\beta}_j$ is then proportional to
\begin{equation}
    f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\bm{\alpha}_j,\bm{\beta}_j).
\end{equation}
We first consider a centering distribution given by a product of independent Gammas:
\begin{equation}
    G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}).
\end{equation}
This model is completed with independent Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.  We also assume a Gamma prior on $\eta$, that is updated via the procedure outlined in \cite{escobar1995}.  We refer to this model as the \emph{projected Gamma--Gamma} (PG--G) model.  An advantage of the PG--G model is that, thanks to conjugacy, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.  Then, the full conditional for $\alpha_{j\ell}$ takes the form
\begin{equation}
    \pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}\times\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\Gamma^{n_j}(\alpha_{j\ell})\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
\end{equation}
for $\ell = 2,\ldots,d$.  For $\ell = 1$, as $\beta_{1} := 1$, the full conditional takes the form
\begin{equation}
    \label{eqn:alpha1update}
    \pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\zeta_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
\end{equation}
Samples of $\alpha_{j\ell}$ can thus be obtained using a Metropolis step. In our analysis, we first transform $\alpha_{j\ell}$ to the log scale, and use a normal proposal density.
The full conditional for $\beta$ is 
\begin{equation}
    \label{eqn:betafc}
    \beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \sim \mathcal{G}\left(\beta_{j\ell}\mid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
\end{equation}
for $\ell = 2,\ldots, d$.  Updating $\beta_{j\ell}$ is done via a Gibbs step.  The hyper-parameters $\xi_{\ell},\tau_{\ell},\zeta_{\ell},\sigma_{\ell}$ follow similar Gamma-Gamma update relationships.  We also explore a restricted form of this model, where $\beta_{\ell} = 1$ for all $\ell$.  Under this model, we use the full conditional in Equation~\ref{eqn:alpha1update} for all $\ell$, and omit inference on $\bm{\zeta},\bm{\sigma}$.  We refer to this model as the \emph{projected restricted Gamma--Gamma} (PRG--G) model.

The second form of centering distribution we explore is a multivariate log-normal distribution on the shape parameters $\bm{\alpha}_j$, with independent Gamma $\beta_{j\ell}$ rate parameters.  
\begin{equation}
    G_0\left(\bm{\alpha}_j,\bm{\beta}_j\mid\bm{\mu},\bm{\Sigma},\zeta,\sigma\right) = \mathcal{LN}\left(\bm{\alpha}_j\mid\bm{\mu},\bm{\Sigma}\right)\times\prod_{\ell = 2}^d\mathcal{G}\left(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}\right).
\end{equation}
This model is completed with a normal prior on $\bm{\mu}$, an inverse Wishart prior on $\bm{\Sigma}$, and Gamma priors on $\zeta_{\ell}$ and $\sigma_{\ell}$, and $\eta$.  This model is denoted as the \emph{projected Gamma--log-normal} (PG--LN) model.  We also explore a restricted Gamma form of this model as above, where $\beta_{\ell} = 1$ for all $\ell$.  This is denoted as the \emph{projected restricted Gamma--log-normal} (PRG--LN) model.  Updates for $\bm{\alpha}$ can be accomplished using a joint Metropolis step, where $\beta_{j\ell}$ for $\ell = 2,\ldots,d$ have been integrated out of the log-density.  That is,
\begin{equation*}
    \begin{aligned}
    \pi(\bm{\alpha}_j\mid\bm{Y},\bm{r},\bm{\delta},\bm{\mu},\bm{\Sigma},\bm{\zeta},\bm{\sigma})
    &\propto
    \frac{1}{\prod_{\ell = 1}^d\alpha_{j\ell}}\exp\left\lbrace-\frac{1}{2}(\log\bm{\alpha}_j - \mu)^T\bm{\Sigma}^{-1}(\log\bm{\alpha}_j - \mu)\right\rbrace \\
    &\hspace{-2cm}\times \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}}{\prod_{\ell = 1}^d\Gamma^{n_j}(\alpha_{j\ell})}\times \prod_{\ell = 2}^d\frac{\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
    \end{aligned}
\end{equation*}
The inferential forms for $\beta_{j\ell}$ and its priors are the same as for PG--G.  The 
  normal prior for $\mu$ is conjugate for the log-normal $\bm{\alpha}_j$, and can be sampled
  via a Gibbs step.  Finally, the inverse Wishart prior for $\Sigma$ is again conjugate to
  the log-normal $\bm{\alpha}_j$, implying that it can also be sampled via a Gibbs step.

### 2022-01-05 results.tex
update

\begin{figure}[htb]
\begin{center}
    \centering
  \caption{Posterior predictive loss (Left) and energy score (right) for various models on simulated data.  Gen indicates baseline.\label{fig:simpples}}
  \begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[height=2in, width=\textwidth]{./images/simulation_ppl}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[height=2in, width=\textwidth]{./images/simulation_es}
  \end{subfigure}
\end{center}
\end{figure}

@@

\begin{figure}[htb]
    \begin{center}
    \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
    \begin{subfigure}[b]{0.48\textwidth}
        \centering 
        \includegraphics[height=2.5in]{./images/erai_grid}
    \end{subfigure}
    %
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \includegraphics[height=2.5in]{./images/era5_grid}
    \end{subfigure}
    \end{center}
\end{figure}

@@

We consider an exploration of the pairwise extremal dependence using Monte Carlo estimates of the 
  coefficients in  Equation~\ref{eqn:chi_ij}. For this we use samples obtained fromn the PRG--G model.
  Figure~\ref{fig:chi_ij} provides a graphical analysis of the results. 
  The coefficients achieve values between $0.286$ and $0.759$ for the ERA-Interim data and 
  between $0.181$ and $0.840$ for the ERA5 data.  The greater range in dependence scores observed 
  with the ERA5 data versus ERA-Interim speaks to the greater granularity of the ERA5 data--distance between locations is a strong contributor to the strength of the 
  pairwise asymptotic dependence. The highest coefficients are $0.759$ for 
  locations 4 and 5 in the ERA-Interim data and
  $0.840$ for locations 1 and 2 in the ERA5 data.  Clearly, pairwise asymptotic
  dependence coefficients tell a limited story, as a particular dependence may include
  more than two locations.   We can, however, glean some information from the patterns that
  emerge in two dimensions.  For the ERA-Interim data, we observe a possible cluster 
  between cells 5-8, indicating a strong dependence among these cells.  Analogously, for
  the ERA5 data, we observe three possible groups of locations.

\begin{figure}[htb]
    \centering
    \caption{Conditional survival curves for selected locations, conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv1d}. The left panel uses original units. Right panel uses standardized units.}
    \includegraphics{./images/condsurv_1d}
\end{figure}

@@

We also explored what effect the power of the projection would have---that is, which $p$-norm
    sphere the data was projected on to for model fitting.  Using the simulated data, coming
    from a mixture of projected Gammas, we were unable to observe an effect from $p$ when 
    looking at a range from 1 to 15 for $p$, on our simulated data.  However, when we repeated 
    this analysis on the IVT data, we were able to observe a drop in the energy score associated 
    with higher $p$, with diminishing returns, as tested up to $p=15$.  The relative performance
    between $p=10$, where the modelling for our presented analysis was conducted, and $p=15$, 
    was not significant.
% EOF

>>>>

\begin{figure}[tb]
\begin{center}
    \centering
  \caption{Posterior predictive loss (Left) and energy score (right) for various models on simulated data.  Gen indicates baseline.\label{fig:simpples}}
  \begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[height=2in, width=\textwidth]{./images/simulation_ppl}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[height=2in, width=\textwidth]{./images/simulation_es}
  \end{subfigure}
\end{center}
\end{figure}

@@

\begin{figure}[tb]
    \begin{center}
    \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
    \begin{subfigure}[b]{0.48\textwidth}
        \centering 
        \includegraphics[height=2.5in]{./images/erai_grid}
    \end{subfigure}
    %
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \includegraphics[height=2.5in]{./images/era5_grid}
    \end{subfigure}
    \end{center}
\end{figure}

@@

We consider an exploration of the pairwise extremal dependence using Monte Carlo estimates of the 
  coefficients in  Equation~\ref{eqn:chi_ij}. For this we use samples obtained from the PRG--G model.
  Figure~\ref{fig:chi_ij} provides a graphical analysis of the results. 
  The coefficients achieve values between $0.286$ and $0.759$ for the ERA-Interim data and 
  between $0.181$ and $0.840$ for the ERA5 data.  The greater range in dependence scores observed 
  with the ERA5 data versus ERA-Interim speaks to the greater granularity of the ERA5 data--distance between locations is a strong contributor to the strength of the 
  pairwise asymptotic dependence. The highest coefficients are $0.759$ for 
  locations 4 and 5 in the ERA-Interim data and
  $0.840$ for locations 1 and 2 in the ERA5 data.  Clearly, pairwise asymptotic
  dependence coefficients tell a limited story, as a particular dependence may include
  more than two locations.   We can, however, glean some information from the patterns that
  emerge in two dimensions.  For the ERA-Interim data, we observe a possible cluster 
  between cells 5-8, indicating a strong dependence among these cells.  Analogously, for
  the ERA5 data, we observe three possible groups of locations.

\begin{figure}[htb]
    \centering
    \caption{Conditional survival curves for selected locations, conditioning on all other dimensions at greater than 90th percentile (fitted)\label{fig:condsurv1d}. The left panel uses original units. Right panel uses standardized units.}
    \includegraphics[width=\linewidth]{./images/condsurv_1d}
\end{figure}

@@

Using our proposed scoring criteria, we explored the effect of the choice of $p$ on the final results. Using the simulated data, generated
    from a mixture of projected Gammas, we were unable to observe sizeable differences in the scores for $p$ 
    ranging between 1 and 15.   However, 
    for the IVT data, we observed a drop in the energy score associated 
    with higher $p$, with diminishing effect as $p$ increased.  
    We observed no significant differences in the performance
    of the model that uses $p=10$, which corresponds to the analysis 
    presented, relative to the one that uses $p=15$.
% EOF
